---
title: "USDA Vegan Foods — Curated AA Dataset"
output: html_notebook
---

This notebook builds a curated subset of the USDA food database for use in the
vegan amino acid LP optimizer. It keeps only plain, consumer-relevant forms of
plant foods with complete essential amino acid (EAA) data, in both dry and
cooked forms where available.

---

## 1 — Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

DATA_DIR <- "~/code/amino_optimize/USDA Data"

fn         <- read.csv(file.path(DATA_DIR, "food_nutrient.csv"))          |> select(fdc_id, nutrient_id, amount)
nutrients  <- read.csv(file.path(DATA_DIR, "nutrient.csv"))                |> select(id, name, unit_name)
food_names <- read.csv(file.path(DATA_DIR, "food_update_log_entry.csv"))  |> select(id, description)
```

---

## 2 — Pivot and label nutrients

```{r pivot}
fn_wide <- fn |>
  pivot_wider(id_cols = fdc_id, names_from = nutrient_id, values_from = amount)

nutrient_lookup <- setNames(
  paste0(nutrients$name, " (", tolower(nutrients$unit_name), ")"),
  as.character(nutrients$id)
)

fn_full <- fn_wide |>
  rename_with(~ nutrient_lookup[.x], .cols = -fdc_id) |>
  left_join(food_names, by = c("fdc_id" = "id")) |>
  select(fdc_id, description, `Protein (g)`, everything())
```

---

## 3 — Vegan base filter (fnf)

Keeps foods with all of:
- `Cholesterol == 0` (vegan proxy; requires the value to be recorded, not NA)
- `Lactose == 0`     (excludes dairy; same caveat — see §5 for plant foods with NA)
- `Protein > 2 g`   (excludes trace-protein foods like spices)
- `Lysine > 0` and `Leucine > 0` (ensures EAA panel was measured)

This yields 273 rows; §4 and §5 narrow these down to ~60 useful entries.

```{r fnf}
fnf <- fn_full |>
  filter(
    !is.na(`Cholesterol (mg)`) & `Cholesterol (mg)` == 0,
    !is.na(`Lactose (g)`)      & `Lactose (g)` == 0,
    !is.na(`Protein (g)`)      & `Protein (g)` > 2,
    !is.na(`Lysine (g)`)       & `Lysine (g)` > 0,
    !is.na(`Leucine (g)`)      & `Leucine (g)` > 0
  ) |>
  arrange(description)

cat("Rows passing vegan + protein filter:", nrow(fnf), "\n")
```

---

## 4 — Curated selection from fnf

Hand-selected to include:
- **Both dry and cooked** forms where available (supports mixed dry/cooked meal plans)
- **No salt**, unfortified, or raw variants preferred over salted/enriched/processed
- **No brands**, restaurant items, junk food, spices, or duplicate preparations

Each kept entry lists the exact USDA description and the reason it was included.
The REMOVED section documents what was dropped and why.

```{r curated}

# ── KEPT ──────────────────────────────────────────────────────────────────────
kept <- tribble(
  ~description,                                                                                    ~note,

  # Grains — dry ---------------------------------------------------------------
  # Dry forms let users model uncooked weight (e.g. measuring out a dry grain mix).

  "Amaranth grain, uncooked",
    "ancient grain; unusually complete EAA for a grain; dry form",

  "Cereals, oats, regular and quick, not fortified, dry",
    "dry rolled oats; not fortified avoids enrichment noise",

  "Cornmeal, degermed, unenriched, yellow",
    "polenta/cornmeal; unenriched; yellow preferred over white (same AAs, clearer identity)",

  "Pasta, dry, unenriched",
    "dry pasta; unenriched variant; enriched adds B vitamins but not relevant to AAs",

  "Rice, brown, long-grain, raw (Includes foods for USDA's Food Distribution Program)",
    "dry brown rice",

  "Rice, white, long-grain, regular, raw, enriched",
    "dry white rice; no 'cooked without salt' entry for plain white rice in this dataset",

  "Rye grain",
    "whole rye berry; kept for completeness alongside rye flour below",

  "Spelt, uncooked",
    "ancient wheat relative; increasingly available in health-food stores",

  "Wheat, KAMUT khorasan, uncooked",
    "ancient grain; higher protein than modern wheat; common in specialty baking",

  # NOTE — the following grains are not in the USDA Data CSVs (food_update_log_entry.csv)
  # and were fetched separately from the USDA API. They are added to fnf_special in §5.
  # Millet raw (169702), Millet cooked (168871)
  # Teff uncooked (169747), Teff cooked (168918)
  # Couscous dry (169699), Couscous cooked (169700)
  # Sorghum grain (169716)
  # Black rice and red rice: no SR Legacy entry in USDA — use brown rice as proxy.

  # Flours (dry) — kept as distinct group for bread/baking page
  "Wheat flour, white (industrial), 11.5% protein, bleached, unenriched",
    "best proxy for all-purpose flour (~11–12% protein); unenriched to avoid vitamin noise; 9/10/13/15% variants dropped as redundant",

  "Rye flour, medium",
    "medium extraction rye flour; representative of common rye flour sold to consumers; light (less fiber) and dark (more bitter) dropped",

  # Grains — cooked ------------------------------------------------------------
  "Cereals, oats, regular and quick, unenriched, cooked with water (includes boiling and microwaving), without salt",
    "cooked oats; without salt",

  "Pasta, cooked, enriched, without added salt",
    "cooked pasta; without salt; enriched/unenriched nutritionally equivalent post-cooking",

  "Pasta, whole-wheat, cooked (Includes foods for USDA's Food Distribution Program)",
    "whole-wheat pasta; higher fiber and slightly better Lys than white pasta",

  "Rice, brown, long-grain, cooked (Includes foods for USDA's Food Distribution Program)",
    "cooked brown rice; without salt",

  "Rice, white, long-grain, parboiled, enriched, cooked",
    "best available proxy for plain cooked white rice in this dataset; parboiling minimally affects AAs",

  # Legumes — dry --------------------------------------------------------------
  "Beans, pinto, mature seeds, raw (Includes foods for USDA's Food Distribution Program)",
    "dry pinto beans; only bean with both dry AND cooked entries with AA data in this dataset",

  "Lentils, raw",
    "dry green/brown lentils; cooked form pulled as special case in §5",

  "Peas, green, split, mature seeds, raw",
    "dry split peas; cooked form not in dataset (only as soup)",

  # Legumes — cooked -----------------------------------------------------------
  "Beans, pinto, mature seeds, cooked, boiled, without salt",
    "cooked pinto beans; without salt",

  "Edamame, frozen, prepared",
    "immature soybeans; high Lys, complete EAA; unique texture vs. mature soy",

  # Nuts — raw/plain -----------------------------------------------------------
  # Raw preferred; roasted/salted variants have nearly identical AAs so dropped.

  "Nuts, almonds",
    "raw almonds",

  "Nuts, brazilnuts, dried, unblanched",
    "exceptional Met+Cys source; LP uses these to satisfy sulfur AA constraint",

  "Nuts, cashew nuts, raw",
    "raw cashews",

  "Nuts, hazelnuts or filberts",
    "raw hazelnuts; blanched and dry-roasted variants dropped",

  "Nuts, macadamia nuts, raw",
    "raw macadamia; lower protein but popular and calorie-dense",

  "Nuts, pecans",
    "raw pecans",

  "Nuts, pine nuts, dried",
    "dried pine nuts",

  "Nuts, pistachio nuts, raw",
    "raw pistachios; relatively high Lys for a nut",

  "Nuts, walnuts, english",
    "raw English walnuts; black walnut variant dropped (less common)",

  # Seeds & spreads ------------------------------------------------------------
  "Seeds, flaxseed",
    "whole flaxseed; ALA omega-3 source; high fiber",

  "Seeds, pumpkin and squash seed kernels, dried",
    "dried pumpkin seed kernels; high His and Ile; roasted variants dropped",

  "Seeds, sesame seed kernels, dried (decorticated)",
    "hulled sesame seeds; good Met source; tahini equivalent in raw form",

  "Seeds, sunflower seed kernels, oil roasted, without salt",
    "roasted sunflower kernels, no salt; only complete-AA entry for sunflower in fnf (dry roasted salted dropped)",

  "Peanut butter, smooth style, without salt",
    "no-salt smooth PB; used as fixed serving in LP (not scalable per 100g)",

  # Vegetables — raw -----------------------------------------------------------
  "Asparagus, raw",
    "modest protein but all 9 EAAs present; popular spring vegetable",

  "Corn, sweet, yellow, raw",
    "raw sweet corn; yellow preferred over white (same AAs)",

  "Kale, raw",
    "raw kale; notably Lys-rich for a leafy green",

  "Mushrooms, oyster, raw",
    "higher protein than button; umami staple in vegan cooking",

  "Mushrooms, portabella, raw",
    "popular meat substitute; raw form",

  "Mushrooms, white, raw",
    "button mushrooms; most common variety",

  "Peas, green, raw",
    "raw green peas",

  "Potatoes, flesh and skin, raw",
    "raw potato; skin included for completeness",

  "Spinach, raw",
    "raw spinach; Lys-rich leafy green",

  # Vegetables — cooked (without salt) ----------------------------------------
  "Broccoli, cooked, boiled, drained, without salt",
    "cooked broccoli; without salt",

  "Corn, sweet, yellow, cooked, boiled, drained, without salt",
    "cooked corn; without salt",

  "Kale, cooked, boiled, drained, without salt",
    "cooked kale; without salt",

  "Mushrooms, shiitake, stir-fried",
    "cooked shiitake; no 'without salt' cooked option in dataset",

  "Peas, green, frozen, cooked, boiled, drained, without salt",
    "frozen green peas cooked; without salt; most common retail form",

  "Potatoes, baked, flesh and skin, without salt",
    "baked potato; without salt",

  "Spinach, frozen, chopped or leaf, cooked, boiled, drained, without salt",
    "cooked frozen spinach; without salt",

  "Sweet potato, cooked, baked in skin, flesh, without salt",
    "baked sweet potato; without salt",

  # Other ----------------------------------------------------------------------
  "Miso",
    "fermented soy paste; unique probiotic staple; high Lys from soy base"
)

# ── REMOVED — rationale (not code) ───────────────────────────────────────────
#
# Brand / restaurant items (~24 entries):
#   RALSTON cereals, MALT-O-MEAL, LUNA BAR, POWER BAR, Burger King hash browns,
#   Carrabba's / Olive Garden spaghetti, Denny's fries, Cracker Barrel fries.
#   → Not consumer pantry staples.
#
# Junk food / snacks (~20 entries):
#   Potato chips, corn chips, crackers (saltines, whole-wheat, cheese-filled),
#   cookies (chocolate chip, sandwich), popcorn (microwave & air), pie, bagel chips,
#   pita chips, SUNCHIPS.
#   → Pass vegan filter (plant oils) but irrelevant for protein planning.
#
# Processed / prepared foods:
#   Garlic bread, English muffins, bread crumbs, egg rolls (vegetable), tortillas
#   (flour, refrigerated and shelf-stable), orange juice, refried beans (canned),
#   baked beans (canned), dehydrated mashed potato (flakes and granules), pancake
#   dry mixes, formulated nutrition bars.
#   → Processed; serving context doesn't suit 100 g LP framework.
#
# Egg white products:
#   "Egg, white, dried" and "Egg, white, raw, frozen, pasteurized".
#   → Not vegan (passes cholesterol=0 filter but should be excluded).
#
# Baby food (2–3 entries):
#   GERBER GRADUATES puffs, Babyfood cereal rice fortified, Babyfood mixed cereal.
#   → Wrong audience; AA values may reflect fortification.
#
# Spices (10+ entries):
#   Basil, chili powder, cinnamon, cloves, curry, garlic powder, ginger, mustard
#   seed, onion powder, oregano, paprika, parsley, black pepper, poppy seed,
#   turmeric.
#   → Protein-dense per 100 g on paper but eaten in teaspoon quantities; including
#     them makes LP solutions unrealistic.
#
# Condiments / sauces:
#   Soy sauce (shoyu, low-sodium shoyu), mustard (prepared yellow), tomato paste.
#   → Condiment quantities; not meaningful protein sources.
#
# Instant / precooked / enriched grain variants:
#   "Rice, white, long-grain, precooked or instant, enriched, dry/prepared",
#   "Rice, white, long-grain, parboiled, unenriched, cooked/dry",
#   "Cereals, oats, instant, fortified, plain, prepared with water",
#   "Cereals, corn grits, white, regular and quick, enriched, dry",
#   "Cereals, farina, enriched, assorted brands including CREAM OF WHEAT".
#   → Instant/precooked forms have similar AAs but are niche; fortified variants
#     have inflated protein numbers from added vitamins.
#
# Duplicate wheat flour variants (kept only 11.5% unenriched):
#   9%, 10%, 13%, 15% protein (enriched and unenriched) — same food, different
#   protein specs for industrial bakeries. Consumer AP flour ≈ 11–12%.
#   The 13% (bread flour) and 15% (high-gluten) are also removed; niche enough
#   to add manually if a bread page needs them.
#
# Rye flour light and dark (kept only medium):
#   Light rye lacks bran (lower protein, lower fiber), dark rye is bitter and
#   niche. Medium is what most home bakers use.
#
# Duplicate nut preparations (kept only raw / plain dried):
#   Almonds (blanched, dry roasted ±salt, oil roasted ±salt),
#   Cashews (oil roasted ±salt),
#   Hazelnuts (blanched, dry roasted),
#   Macadamia (dry roasted),
#   Pecans (dry roasted ±salt, oil roasted ±salt),
#   Pistachios (dry roasted ±salt),
#   Walnuts, black (dried) — kept English walnut instead.
#   Almond butter, sunflower seed butter — kept whole-nut / whole-seed forms.
#   Mixed nuts (various) — too variable to represent a single food.
#   → Roasting has minimal effect on AAs. Salt is a preparation choice, not a
#     nutritional one. Raw avoids any ambiguity.
#
# Salted / canned legume variants:
#   Beans pinto (with salt, canned drained), kidney (canned ×4 variants),
#   navy (with salt), beans baked (canned vegetarian).
#   → Same nutrition as without-salt; canned adds sodium not relevant to LP.
#
# Duplicate corn variants:
#   White sweet corn (cooked ±salt, canned), yellow corn (canned vacuum pack ×3,
#   frozen on-cob ×4, frozen kernels ±salt, microwaved).
#   → Kept raw and one cooked-without-salt; all others redundant.
#
# Asparagus with salt, broccoli with salt, kale with salt/frozen,
# spinach with salt/frozen with salt, sweet potato with salt,
# peas green (canned, with salt, frozen with salt):
#   → Salted or canned duplicates of kept entries.
#
# Potatoes processed:
#   French fries (oven, frozen, shoestring, crinkle, steak — all brands),
#   hash browns (frozen pan-fried, home-prepared), red/Russet/white flesh baked.
#   → Processed; skin-off variants give incomplete nutrition vs. flesh+skin kept.
#
# Duplicate peanut butter (kept smooth without salt):
#   Chunk style with/without salt, smooth with salt, "Peanut Butter smooth
#   (Includes foods for USDA's Food Distribution Program)".
#   → Same food; without-salt smooth is the cleanest entry.
#
# Peanuts whole (kept PB):
#   "Peanuts, all types, dry-roasted ±salt", "oil-roasted ±salt".
#   → PB is the dominant consumer form; whole peanuts can be added manually.
#
# Miscellaneous:
#   Alfalfa seeds sprouted raw — niche; minimal protein contribution.
#   Dried fruits (apricots, dates, figs, prunes) — very low protein per 100 g
#     once sugar content is accounted for; not meaningful for LP.
#   Broccoli raab — less common; covered by regular broccoli.
#   Brussels sprouts raw — marginal protein; no distinct AA advantage over kale.
#   Lima beans (immature, frozen, ±salt) — similar profile to pinto/navy.
#   Cream substitute powdered — not vegan in spirit (dairy-adjacent flavor).
#   Restaurant spaghetti entries (Italian, Mexican rice) — not reproducible.
#   Egg rolls vegetable frozen — processed, minor protein.

# ── Filter fnf to kept descriptions ──────────────────────────────────────────
fnf_curated <- fnf |>
  filter(description %in% kept$description) |>
  left_join(kept |> select(description, note), by = "description") |>
  arrange(description)

cat("Rows from fnf in curated list:", nrow(fnf_curated), "\n")

# Verify all kept descriptions matched (catch typos)
unmatched <- kept$description[!kept$description %in% fnf_curated$description]
if (length(unmatched) > 0) {
  cat("\n*** UNMATCHED descriptions (check spelling):\n")
  walk(unmatched, ~ cat(" -", .x, "\n"))
}
```

---

## 5 — Special cases: plant foods with Lactose = NA

Several plant foods fail the strict `Lactose == 0` filter because their lactose
column is `NA` (data not recorded) rather than `0`. These are obviously dairy-free;
we pull them directly from `fn_full` and document each one.

```{r special_cases}

# Each entry: fdc_id, reason Lactose = NA, brief note.
special_ids <- tribble(
  ~fdc_id,  ~note,

  # Grains / flours
  170687L,  "Buckwheat flour, whole-groat — Lactose NA; plant food, no dairy; protein 12.6 g, full AA panel; best dry flour for buckwheat pancakes / soba",

  # Legumes — cooked
  172421L,  "Lentils, mature seeds, cooked, boiled, without salt — Lactose NA; plant food; cooked counterpart to dry lentils kept in fnf_curated",

  # Legumes — dry (additional variety)
  174284L,  "Lentils, pink or red, raw — Lactose NA; plant food; higher protein (23.9 g) than green lentils; popular in South Asian dal and quick-cooking recipes",

  # Protein supplement
  173181L,  "Beverages, Protein powder soy based — Lactose NA; plant food; protein 55.6 g, Lys 3.55 g, Leu 4.60 g; useful as soy-based protein powder entry for supplement page",

  # Grains added via USDA API (not present in food_update_log_entry.csv)
  169702L,  "Millet, raw — dry millet; protein 11.0 g; full AA panel",
  168871L,  "Millet, cooked — cooked millet; without salt",
  169747L,  "Teff, uncooked — dry teff; protein 13.3 g; notably high Met; Ethiopian staple",
  168918L,  "Teff, cooked — cooked teff; without salt",
  169699L,  "Couscous, dry — semolina-based; dry form; similar AA profile to pasta",
  169700L,  "Couscous, cooked — cooked couscous; without salt",
  169716L,  "Sorghum grain — dry whole grain; gluten-free; protein 10.6 g"
  # Black rice and red rice: no USDA SR Legacy entry — brown rice is the best proxy.
)

fnf_special <- fn_full |>
  filter(fdc_id %in% special_ids$fdc_id) |>
  left_join(special_ids, by = "fdc_id") |>
  select(fdc_id, description, `Protein (g)`, `Lysine (g)`, `Leucine (g)`,
         `Cholesterol (mg)`, `Lactose (g)`, note) |>
  arrange(description)

cat("Special-case rows:", nrow(fnf_special), "\n")
print(fnf_special |> select(fdc_id, description, `Protein (g)`, note))
```

---

## 6 — Combined final dataset

```{r final}
# Core columns for AA work (extend as needed)
core_cols <- c("fdc_id", "description", "Protein (g)",
               "Lysine (g)", "Leucine (g)", "Isoleucine (g)",
               "Valine (g)", "Methionine (g)", "Cystine (g)",
               "Phenylalanine (g)", "Tyrosine (g)",
               "Threonine (g)", "Tryptophan (g)", "Histidine (g)",
               "note")

safe_select <- function(df) {
  df |> select(any_of(core_cols))
}

fnf_final <- bind_rows(
  fnf_curated  |> safe_select(),
  fnf_special  |> safe_select()
) |>
  arrange(description)

cat("\nFinal curated dataset:", nrow(fnf_final), "foods\n")
View(fnf_final)
```

---

## 7 — Chickpea cultivar note

The LP uses USDA fdc 173757 (chickpeas cooked, from the Soy_Cube dataset via
`build_foods.py`). A comparison with published cultivar data shows:

| AA          | USDA % of protein | Desi NIFA-88 | Kabuli Hassan-2k |
|-------------|:-----------------:|:------------:|:----------------:|
| Lys         | 6.69              | 6.7 ✓        | 7.8              |
| **Leu**     | **7.12**          | **8.2**      | **9.5**          |
| Thr         | 3.71              | 2.7          | 3.5              |
| Trp         | 0.96              | 0.6          | 1.1              |
| Val         | 4.20              | 4.1 ✓        | 5.2              |
| Met+Cys     | 2.65              | 1.2          | 2.1              |
| Phe+Tyr     | 7.84              | 7.6 ✓        | 9.3              |

US supermarket chickpeas are predominantly **Kabuli** (large, beige, round).
However, the USDA SR Legacy entry likely reflects averaged assay data across
available samples — not a single cultivar. Using USDA values as-is is the most
defensible choice for a US-focused site. If Leu turns out to be a binding
constraint in LP solutions, the USDA value (7.12%) may be slightly conservative.

```{r chickpea_note, echo=FALSE}
cat("Chickpea USDA fdc 173757 (cooked, from Soy_Cube dataset) retained as-is.\n")
cat("Cultivar note added above for transparency.\n")
```
